Correctness of Algebraic File Synchronization

### 2015-11-22 2015/paper.tex
Describe ordering in detail

\newcommand{\setv}{\mathcal{V}} % Set of values
\newcommand{\setvx}[1]{\mathcal{V}_{#1}}
\newcommand{\setf}{\setvx{F}} % Set of file values
\newcommand{\setd}{\setvx{D}} % Set of directory values
\newcommand{\setb}{\setvx{\empt}} % Set of empty values
\newcommand{\setn}{\mathcal{N}} % Set of nodes / paths
\newcommand{\empt}{\bot}
\newcommand{\parent}{\mathtt{parent}}
\newcommand{\topnode}{\mathtt{None}} % return value of parent()
\newcommand{\fsbroken}{\mathtt{Broken}} % broken FS value
\newcommand{\FS}{\Phi} % a filesystem
\newcommand{\GS}{\Xi} % another filesystem
\newcommand{\pp}{p^*} % parent path

@@

% Insight: (Prof. Norman Ramsey)
% - updates where two orders are not equivalent are incompatible
% - we only have a set of updates and need to create an order

@@

The following concepts will allow us to investigate sequences independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences of commands.
The first two relations echo the ones defined in \cite{NREC:alg}.

@@

% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{myax}\label{ax_distantchild_breaks}
$\cxy(\pp,V_Y)\cc  \czw(p,V_W) \equiv \cbrk$ where $\pp\descendant p$ and $\pp\neq\parent(p)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$.
\end{myax}

\begin{myax}\label{ax_distantparent_breaks}
$\cxy(p,V_Y)\cc  \czw(\pp,V_W) \equiv \cbrk$ where $p\ancestor \pp$ and $\parent(p)\neq \pp$ and $\cxy\neq\cbb$ and $\czw\neq\cdd$.
TODO merge into previous
\end{myax}

\begin{mydef}{Construction pair.}
A pair of commands on paths $\pp$ and $p$ is a construction pair if $\pp=\parent(p)$ and the commands are one
of the following:
   \begin{gather*}
   % bd>bf
            [\cbd(\pp, D)\cc  \cbf(p, F)] \\
   % bd>bd
            [\cbd(\pp, D_1)\cc  \cbd(p, D_2)] \\
   % fd>bf
            [\cfd(\pp, D)\cc  \cbf(p, F)] \\
   % fd>bd
            [\cfd(\pp, D_1)\cc  \cbd(p, D_2)]
   \end{gather*}
\end{mydef}

\begin{mydef}{Destruction pair.}
A pair of commands on paths $p$ and $\pp$ is a destruction pair if $\parent(p)=\pp$ and the commands are one
of the following:
   \begin{gather*}
   % fb<db ??
            [\cfb(p)\cc  \cdb(\pp)] \\
   % fb<df ??
            [\cfb(p)\cc  \cdf(\pp, F)] \\
   % db<db ??
            [\cdb(p)\cc  \cdb(\pp)] \\
   % db<df ??
            [\cdb(p)\cc  \cdf(\pp, F)]
   \end{gather*}
\end{mydef}

\begin{myax}\label{ax_directchild_breaks}
$\cxy(\pp,V_Y)\cc  \czw(p,V_W) \equiv \cbrk$ where $\pp=\parent(p)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$
and the pair is not a construction pair.
\end{myax}

\begin{myax}\label{ax_directparent_breaks}
$\cxy(p,V_Y)\cc  \czw(\pp,V_W) \equiv \cbrk$ where $\parent(p)=\pp$ and $\cxy\neq\cbb$ and $\czw\neq\cdd$
and the pair is not a destruction pair.
\end{myax}

\begin{myax}\label{ax_child_assert}
An assertion command can be added on a child path:
$\cbb(p)\cc \cxy(\pp,V_Y) \equiv \cxy(\pp,V_Y) \equiv \cxy(\pp,V_Y)\cc \cbb(p)$ where $\pp\descendant p$ and $\cxy\neq\cdd$.
\end{myax}

\begin{myax}\label{ax_parent_assert}
An assertion command can be added on the parent path:
$\cdd(\pp)\cc \cxy(p,V_Y) \equiv \cxy(p,V_Y) \equiv \cxy(p,V_Y)\cc \cdd(\pp)$ where $\pp\descendant p$ and $\cxy\neq\cbb$.
\end{myax}

@@

\begin{mydef}{Minimal set or sequence.}
A sequence or set of commands is minimal if it contains at most one command on every node.
\end{mydef}

@@

\begin{myax}\label{ax_assert}
$\cxy(p,V_Y) \eqext \emptyseq$ where $\cxy=\cbb$ or $\cxy=\cdd$.
\end{myax}

@@

We often encounter the case where we only have a set of commands without a specified order.
Indeed, it is actually the task of the reconciler to determine whether there is an order
in which updates from multiple replicas can be applied to a filesystem.

If $U$ is a minimal set of commands
that we know can be applied to at least one filesystem in some order without breaking it,
then we can split $U$ into a maximum number of disjunct subsets
(components) so that the paths of two commands from two different subsets are never directly related.
If we aim to order the commands in $U$, then we can allow any permutation of 
the components as the commands in them will commute freely.

Components containing a single command do not pose a problem when determining
possible orders.
(Note that $\cff$ can only appear in a component of one.)

In each component that is larger than one, the paths 
form a rooted ordered tree via the parent--child relationships,
and the pairs connected by edges will be like pairs in the
construction or destruction laws.
To order the commands in a component,
select the command with the topmost path, so that there would be no command
on its parent path (if there is one). 
If the command removes content at this path (it is of the form $\fsregcommand{X}{\empt}$), then it
needs to follow all other commands in the component, and
the component is made of destruction pairs.
Otherwise the command with the topmost component needs to precede other commands
in the component, and it is made of construction pairs.
Move the command with the topmost path in place, and
repeat the process of splitting commands into subsets on the remaining commands
in the component. Recursively, this way we can determine all allowed orders of the commands.

\begin{mydef}{$\orderset{U}$}
We use $\orderset{U}$ to denote the set of sequences that contain all commands in $U$
and represent all orders generated by this algorithm.
By $\ordered{U}$ we mean an arbitrary sequence in the set.
\end{mydef}

As a specific case, we can regard a sequence of commands $S$
as a set and determine its $\orderset{S}$.

\begin{mylem}\label{lemma:neighbour}
If $U$ contains commands on both $p$ and $\pp$, then
there is a $U'\in\orderset{U}$ where they are next to each other.
\end{mylem}

\begin{mydef}
$\extset{S} := \{S'|S\eqext S'\}$
\end{mydef}

\begin{mylem}\label{lemma:minextset}
For any minimal sequence $S$ not equivalent to $\cbrk$, $\extset{S} = \{S'|S\equiv S'\}$.
\end{mylem}
\begin{proof}
This is because as $S$ has at most one command on each path, only equivalence axioms
can be applied to it.
\end{proof}

\begin{mylem}\label{lemma:minextorder}
For any minimal sequence $S$ not equivalent to $\cbrk$,
$\orderset{S} = \extset{S}$.
\end{mylem}

% TODO Can these be derived from the algebra?

\section{Soundness}

\section{Completeness}

\begin{mylem}
\((\forall\FS: S\FS = \empt) \Rightarrow S \equiv \cbrk\)
\end{mylem}

@@

If $B$ has a command on $q$, then
from \cref{lemma:neighbour}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxy$,
and from \cref{lemma:minextset,lemma:minextorder} we know that $B'\equiv B$.
% TODO Use when describing ordering:
Considering the eight construction and destruction pairs we see that the command
on the parent path determines which group the pair belongs to.
This in turn determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $q$ must be on
the same side of the command on $p$ in both sequences.

@@

As $S$ is simple and $\works{S}$, from 
\cref{ax_distantchild_breaks,ax_distantparent_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $S$,
with the number on relevant commands not in $S^*$ reduced by one.
In the last two cases,
either $q=\parent(p)$ or $p=\parent(q)$.

>>>>

\newcommand{\setv}{\mathcal{V}} % Set of values
\newcommand{\setvx}[1]{\mathcal{V}_{#1}}
\newcommand{\setf}{\setvx{F}} % Set of file values
\newcommand{\setd}{\setvx{D}} % Set of directory values
\newcommand{\setb}{\setvx{\empt}} % Set of empty values
\newcommand{\setn}{\mathcal{N}} % Set of nodes / paths
\newcommand{\empt}{\bot}
\newcommand{\parent}{\mathtt{parent}}
\newcommand{\topnode}{\mathtt{None}} % return value of parent()
\newcommand{\fsbroken}{\mathtt{Broken}} % broken FS value
\newcommand{\FS}{\Phi} % a filesystem
\newcommand{\GS}{\Xi} % another filesystem
\newcommand{\nn}{n^*} % parent node

@@

% ORDERING - because when taking updates from different replicas we get an essentially unordered set
% Insight: (Prof. Norman Ramsey)
% - updates where two orders (AB vs BA) are not equivalent are incompatible
% - we only have a set of updates and need to create an order

@@

The following concepts allow us to investigate sequences independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences of commands.
These relations echo the ones defined in \cite{NREC:alg}.

@@

\begin{myax}\label{ax_distantrel_breaks}
Commands on distant relatives break all filesystems:
$\cxy(\nn,V_Y)\cc  \czw(n,V_W) \equiv \cbrk$
and $\czw(n,V_W)\cc \cxy(\nn,V_Y) \equiv\cbrk$
where $\nn\descendant n$ and $\nn\neq\parent(n)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$.
\end{myax}

\begin{mydef}{Construction pair.}
A pair of commands on nodes $\nn$ and $n$ is a construction pair if $\nn=\parent(n)$ and the commands are one
of the following:
   \begin{gather*}
   % bd>bf
            \cbd(\nn, D)\cc  \cbf(n, F) \\
   % bd>bd
            \cbd(\nn, D_1)\cc  \cbd(n, D_2) \\
   % fd>bf
            \cfd(\nn, D)\cc  \cbf(n, F) \\
   % fd>bd
            \cfd(\nn, D_1)\cc  \cbd(n, D_2)
   \end{gather*}
\end{mydef}

\begin{myax}\label{ax_directchild_breaks}
All other commands on a child break every filesystem:
$\cxy(\nn,V_Y)\cc  \czw(n,V_W) \equiv \cbrk$ where $\nn=\parent(n)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$
and the pair is not a construction pair.
\end{myax}

\begin{mydef}{Destruction pair.}
A pair of commands on nodes $n$ and $\nn$ is a destruction pair if $\parent(n)=\nn$ and the commands are one
of the following:
   \begin{gather*}
   % fb<db ??
            \cfb(n)\cc  \cdb(\nn) \\
   % fb<df ??
            \cfb(n)\cc  \cdf(\nn, F) \\
   % db<db ??
            \cdb(n)\cc  \cdb(\nn) \\
   % db<df ??
            \cdb(n)\cc  \cdf(\nn, F)
   \end{gather*}
\end{mydef}

\begin{myax}\label{ax_directparent_breaks}
All other commands on a parent break every filsystem:
$\cxy(n,V_Y)\cc  \czw(\nn,V_W) \equiv \cbrk$ where $\parent(n)=\nn$ and $\cxy\neq\cbb$ and $\czw\neq\cdd$
and the pair is not a destruction pair.
\end{myax}

\begin{myax}\label{ax_child_assert}
An assertion command can be added on a descendant node:
$\cbb(n)\cc \cxy(\nn,V_Y) \equiv \cxy(\nn,V_Y) \equiv \cxy(\nn,V_Y)\cc \cbb(n)$ where $\nn\descendant n$ and $\cxy\neq\cdd$.
\end{myax}

\begin{myax}\label{ax_parent_assert}
An assertion command can be added on an ancestor node:
$\cdd(\nn)\cc \cxy(n,V_Y) \equiv \cxy(n,V_Y) \equiv \cxy(n,V_Y)\cc \cdd(\nn)$ where $\nn\descendant n$ and $\cxy\neq\cbb$.
\end{myax}

\begin{myax}\label{ax_assert}
Assertion commands can be removed:
$\cxy(n,V_Y) \eqext \emptyseq$ where $\cxy=\cbb$ or $\cxy=\cdd$.
\end{myax}

% Update detection
% ----------------

In a command-based reconciliation solution we assume that we have two sequences of commands
$A$ and $B$ that have modified a single filesystem $\FS$ yielding two different replicas $\FS_A$ and $\FS_B$ that we
need to reconcile. While it is conceivable that the sequences would be based on a record of
all operations that modified the two filesystems, in most filesystem implementations
such records are not implemented, and therefore we must construct suitable sequences
by comparing $\FS_A$ and $\FS_B$ to their common ancestor, $\FS$. This is called update detection.

A sequence that transforms $\FS$ into $\FS_A$ must contain at least one command on each node
that changes between the two states of the filesystem. Also, it is easy to see that this is also
sufficient, as there are commands available for any input and output value. 
Specifically, if at node $n$, $\FS(n)=X'$ and $\FS_A(n)=Y'\neq X'$, then we add the command $\cxy(n,Y')$ to $A$
where $X'\in\setvx{X}$ and $Y'\in\setvx{Y}$.

To be able to describe such sequences better, let us introduce the following properties:

\begin{mydef}{Minimal set or sequence.}
% TODO Don't introduce?
A sequence or set of commands is minimal if it contains at most one command on every node.
\end{mydef}

This update detector therefore yields a simple set of commands because we only add a single command
for each node, and we only add commands that are necessary, that is, there will be no 
assertion commands in the set.

The next step in generating the sequences is to order the commands collected.
As this task is at the heart of reconciliation itself independently of update detection,
we discuss it in the next section.

% TODO Proof of correctness of update detection - the generated order will work

% Ordering commands
% ----------------

We often encounter the case where we only have a set of commands without a specified order.
As we have seen above, this can occur after the first stage of update detection,
but, more importantly,
it is actually the task of the reconciler to determine whether there is an order,
and if yes, what order,
in which updates from multiple replicas can be applied to a filesystem.

In this section we describe an algorithm with which all possible orderings of a
set of sequences can be generated, but before doing so, let us consider the following lemma.

\begin{mylem}\label{connected_changes}
If a simple set of commands that can be applied to a filesystem in some order without breaking it
contains commands on $n$ and $\nn$ where $\nn\descendant n$
but $\nn\neq\parent(n)$, then the set must also contain a command on $\parent(n)$.
\end{mylem}
\begin{proof}
We will prove this statement for simple sequences, as if all sequences must contain a command on $\parent(n)$
under the given conditions, then so must sets because otherwise there would be no order in which thay can be
applied to any filesystem.

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxy(n,V_Y)$ on $n$
and another command $\czw(\nn,V_W)$ on $\nn$.
We use an indirect proof and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxy$ and $\czw$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxy(n,V_Y)$ in the direction where $\czw(\nn,V_W)$ is.
Let this command be $\cqr(m,V_R)$.
If $\czw$ is to the right, the $A$ looks like the following:
\[ A = \cdots\cc\cxy(n,V_Y)\cc\cqr(m,V_R)\cc\cdots\cc\czw(\nn,V_W)\cc\cdots \]
If $n\unrel m$, reverse $\cxy$ and $\cqr$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contains assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxy$ and $\czw$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

By applying \cref{connected_changes} multiple times we see that if a simple set of commands
contains two commands on related nodes, then it must contain a command on each
node that is between these two nodes in the parent--child tree.

We can now define our algorithm to generate orders of sets of commands.

Let $U$ be a simple set of commands
that we know can be applied to at least one filesystem in some order without breaking it.
Then we can split $U$ into a maximum number of disjunct subsets
(components) so that the nodes of two commands from two different subsets are unrelated.
These will simply be the components of the subset of the node forest that contains the nodes affected by the set.

From \cref{ax_separate_commute} we know that commands from different 
components can be applied in any order.
Therefore once we have ordered the commands inside a component, we generate
all permutations of the components themselves, including the ones
in which commands from separate components mix and overlap
(but where the possible orders determined for the components are respected).

Next, consider the ordering inside a component.

This is not a problem in the case of components containing a single command only.
(Note that $\cff$ can only appear in a component of one.)

In each component that is larger than one, 
we know that all nodes are related, and from \cref{connected_changes}
we also know that it must contain commands on all nodes in between.
Therefore the nodes
form a rooted ordered tree,
and the pairs of commands connected by the edges in the tree 
must be construction or destruction pairs.
(Otherwise, there would be no ordering of the component that is not equivalent to $\cbrk$
and via \cref{ax_separate_commute} there would be no ordering of the whole set, either.)

To order the commands in the component,
select the command with the topmost node, so that there would be no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\fsregcommand{X}{\empt}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any sequence to break all filesystems.

Then move the command with the topmost node to the beginning or the end of
the sequence we are creating out of the component,
and recursively start the whole algorithm again on the commands in the component that remain.
The result of this process is a set of possible orders for each component,
which are then combined into a set of possible orders for the whole set of commands.

\begin{mydef}{$\orderset{U}$}
We use $\orderset{U}$ to denote the set of sequences that contain all commands in
the simple set $U$
and represent all orders generated by this algorithm.
\end{mydef}

As a specific case, we can regard a sequence of commands $S$
as a set and determine its $\orderset{S}$.
Let us investigate the properties of simple sequences and this set a bit further.

\begin{mydef}{$\extset{S}$.}
For any sequence $S$ we can consider all sequences that extend it:
$\extset{S} := \{S'|S\eqext S'\}$
\end{mydef}

% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{mylem}\label{simple_is_max}
For any simple sequence $S\neq\cbrk$, $\forall T: S\eqext T \Rightarrow S\equiv T$;
that is, the domains of simple sequences cannot be extended further.
This is equivalent to $\extset{S} = \{S'|S\equiv S'\}$.
\end{mylem}
\begin{proof}
TODO
% TODO
\end{proof}

\begin{mylem}\label{lemma:minextorder}
% TODO orderset only defined on simple
For any minimal sequence $S\neq\cbrk$, $\orderset{S} = \extset{S}$.
\end{mylem}
\begin{proof}
TODO
% TODO
\end{proof}

\begin{mylem}
All sequences in $\orderset{S}$ of a simple sequence $S$ are equivalent.
\end{mylem}
\begin{proof}
This follows from \cref{simple_is_max,lemma:minextorder}.
\end{proof}

TODO: we'll treat sequences as sets and sets as sequences
once we know they can be applied.

\begin{mylem}\label{lemma:neighbor}
If $U$ contains commands on both $n$ and $\parent(n)$, then
there is a $U'\in\orderset{U}$ where they are next to each other.
\end{mylem}

@@

@@

@@

@@

If $B$ has a command on $q$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxy$,
and from \cref{lemma:minextset,lemma:minextorder} we know that $B'\equiv B$.
% TODO Use when describing ordering:
Considering the eight construction and destruction pairs we see that the command
on the parent path determines which group the pair belongs to.
This in turn determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $q$ must be on
the same side of the command on $p$ in both sequences.

@@

As $S$ is simple and $\works{S}$, from 
\cref{ax_distantrel_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $S$,
with the number on relevant commands not in $S^*$ reduced by one.
In the last two cases,
either $q=\parent(p)$ or $p=\parent(q)$.
